Closing the gap between statistical and scientific workflows for improved forecasts in ecology

### 2025-04-01 forecastflows.tex
add some refs

% There is thus room for improvment, with a workflow that addresses theses flaws while taking account the reality of working with ecological data.
% Moving along the data-model space, step-by-step
% \item stressing the need to think about model before study design
% \item 
To address these flaws while accounting for the realities of working with ecological data, a comprehensive workflow is needed to move along the data-model space, following a coherent sequence of steps. In particular, more efforts should be placed on model evaluation before incorporating any real data. This would force the modeler to acknowledge that some parameters might be non-identifiable and to reconsider the model structure. Similarly, it is essential to assess whether model predictions---once parameters are informed by data---are consistent with observations. The strength of such workflow lies in its flexibility, making it applicable to a wide range of modeling approaches, from simple trend analyses to more complex process-based models. At each step, the modeler need to critically examine its understanding of ecological processes, questioning previous assumptions, and explicitly acknowledge sources of uncertainty. This approach has the potential to enhance model interpretability and allow for a more transparent evaluation of model strengths and limitations. It also replace parameters at the core of the modeling process, as fundamental components that shape both inference and forecasting.

% More details on the workflow:  walk through the different steps
% the paragraph is a bit loooong
The first step of the workflow is to define an explicit research question and formulate some hypotheses. This involves making assumptions about the most influential drivers, within the specific context of our study. This step help us to think about the mechanisms that could generate the data we observe, including the observational error. Naturally, this leads to the development of a model---an ensemble of mathematical equations that encapsulates our knowledge and designed to answer our research question. The general idea is to start with a relatively simple model, that we could refine later. At this stage, prioritizing biologically meaningful parameters is crucial, as it allows us to have a sense of plausible parameter values. The next step is to simulate data that reflect our assumptions about the data structure and biases, by fixing parameter values to some values (which is straightforward if the parameters are interpretable). We then fit our model to this simulated dataset and evaluate its ability to recover the prescribed values. Some issues might arise at this step. Structural degeneracies in the model might become apparent, revealing that some parameters are highly non-identifiable. This is the first feedback loop: if the model is not able to recover known parameter values, we must reconsider its structure. This way, we ensure the model works well on simulated data before incorporating real observations. We can then fit it to real data, to obtain parameter estimates constrained by observations. Here, difficulty in fitting the model might indicate an inherent need for more data to address our initial question. This could lead us to either simplify our research question or---ideally---launch new data collection efforts. The next step is a new data simulation step, this time using our fitted model to generate predictions. This retrodictive check allows model output to be compared to observations. Discrepancies may indicate a missing key driver---perhaps an expected outcome if we known our initial model was too simplistic. This is a second feedback loop. We can refine the model to integrate the missing process and restart the workflow. Insights from the retrodictive check can also lead us to introduce additional complexity when simulating fake input data, such as phylogenetic structure or observational biases (e.g. unbalanced data). This iterative evaluation of the model moves beyond a simple reliance on goodness-of-fit metrics. At each iteration, we are able to evaluate the model behavior, both with simulated and real data, taking into account our expert knowledge of the ecological processes. Within such a workflow, forecasting emerges as a natural outcome: rather than being a final goal, it only involves jointly modeling new circumstances along with the original data.

@@

% Forecasting - outline current problem
Ecological forecasting is a broad field with a diverse range of methods. Process-based modeling is often considered the gold standard in ecology \citep{Urban2016, Pilowsky2022} and beyond. Process-based models are built on explicit mathematical equations to describe (supposedly causal) relationships between environmental drivers and ecological responses. They also often incorporate empirical relationships, particularly when knowledge is incomplete or when some processes are intentionally omitted. Processes are often represented at different nested spatiotemporal scales, depending on the underlying assumptions. Model development is the central step, typically requiring several years, yet it often remains opaque from an external perspective. The step of designing the model---translating knowledge and hypotheses into mathematical equations and parameters---is often blurred with the step of model calibration (or tuning), where parameter values are inferred. Models are often treated as an accumulation of multiple submodels, each governing one or several ecological processes. Rather than being fitted as a whole, submodels are calibrated separately against specific subsets of data, and some parameters are simply prescribed (i.e. fix to a value found in the literature) or tuned to reproduce some observations or theory. The way models are currently calibrated is not accidental, but rather a constraint imposed by their structure, where many parameters compensate for one another.

These limitations are central to current debates. In climate modeling, increasing model complexity has not necessarily led to reduced uncertainty. For instance, the uncertainty range on the effect of increasing CO2 concentration on temperature have remained largely unchanged \citep{}. This has driven calls for more rigorous and transparent calibration processes \citep{balaji2022general}. Similar concerns arise in ecology, where strong discrepancies exist between model projections \citep{}. A promising avenue is the simplification of models, avoiding over-parametrization when the data provide little information to constrain some parameters \citep{Wang2017, Harrison2021}. If a model becomes too complex, understanding the sources of uncertainty and how they propagate through the model may become nearly impossible.
Each additional process and parameter can increase overall uncertainty to the point where predictions lose their usefulness for decision makers \citep{Saltelli2020}.

>>>>

% There is thus room for improvment, with a workflow that addresses theses flaws while taking account the reality of working with ecological data.
% Moving along the data-model space, step-by-step
% \item stressing the need to think about model before study design
% \item 
To address these flaws while accounting for the realities of working with ecological data, a comprehensive workflow is needed to move along the data-model space, following a coherent sequence of steps (Fig 1). In particular, more efforts should be placed on model evaluation before incorporating any real data. This would force the modeler to acknowledge that some parameters might be non-identifiable and to reconsider the model structure. Similarly, it is essential to assess whether model predictions---once parameters are informed by data---are consistent with observations. The strength of such workflow lies in its flexibility, making it applicable to a wide range of modeling approaches, from simple trend analyses to more complex process-based models. At each step, the modeler need to critically examine its understanding of ecological processes, questioning previous assumptions, and explicitly acknowledge sources of uncertainty. This approach has the potential to enhance model interpretability and allow for a more transparent evaluation of model strengths and limitations. It also replace parameters at the core of the modeling process, as fundamental components that shape both inference and forecasting. % I think we would need to explain more this idea about parameters... of keep it for later

% More details on the workflow:  walk through the different steps
% the paragraph is a bit loooong
The first step of the workflow is to define an explicit research question and formulate some hypotheses (Fig 2). This involves making assumptions about the most influential drivers, within the specific context of our study. This step help us to think about the mechanisms that could generate the data we observe, including the observational error. Naturally, this leads to the development of a model---an ensemble of mathematical equations that encapsulates our knowledge and designed to answer our research question. The general idea is to start with a relatively simple model, that we could refine later. At this stage, prioritizing biologically meaningful parameters is crucial, as it allows us to have a sense of plausible parameter values. The next step is to simulate data that reflect our assumptions about the data structure and biases, by fixing parameter values to some values (which is straightforward if the parameters are interpretable). We then fit our model to this simulated dataset and evaluate its ability to recover the prescribed values. 
% Some issues might arise at this step
Structural degeneracies in the model might become apparent, revealing that some parameters are highly non-identifiable. This is the first feedback loop: if the model is not able to recover known parameter values, we must reconsider its structure. This way, we ensure the model works well on simulated data before incorporating real observations. We can then fit the model to real data, to obtain parameter estimates constrained by observations. Here, difficulty in fitting the model might indicate an inherent need for more data to address our initial question. This could lead us to either simplify our research question or---ideally---launch new data collection efforts. The next step is a new data simulation step, this time using our fitted model to generate predictions. This retrodictive check allows model output to be compared to observations. Discrepancies may indicate a missing key driver---perhaps an expected outcome if we known our initial model was too simplistic. This is a second feedback loop. We can refine the model to integrate the missing process and restart the workflow. Insights from the retrodictive check can also lead us to introduce additional complexity when simulating fake data, such as phylogenetic structure or observational biases (e.g. unbalanced data). This iterative evaluation of the model moves beyond a simple reliance on goodness-of-fit metrics. At each iteration, we are able to evaluate the model behavior, both with simulated and real data, taking into account our expert knowledge of the ecological processes. Within such a workflow, forecasting emerges as a natural outcome: rather than being a final goal, it only involves jointly modeling new circumstances along with the original data.

@@

% Forecasting - outline current problem
Ecological forecasting is a broad field with a diverse range of methods. Process-based modeling is often considered the gold standard in ecology \citep{Urban2016, Pilowsky2022} and beyond. Process-based models are built on explicit mathematical equations to describe (supposedly causal) relationships between environmental drivers and ecological responses. They also often incorporate empirical relationships, particularly when knowledge is incomplete or when some processes are intentionally omitted. Processes are often represented at different nested spatiotemporal scales, depending on the underlying assumptions. Model development is the central step, typically requiring several years, yet it often remains opaque from an external perspective. The step of designing the model---translating knowledge and hypotheses into mathematical equations and parameters---is often blurred with the step of model calibration (or tuning), where parameter values are inferred. Models are often treated as an accumulation of multiple submodels, each governing one or several ecological processes. Rather than being fitted as a whole, submodels are calibrated separately against specific subsets of data, and some parameters are simply prescribed (i.e. fix to a value found in the literature) or tuned to reproduce some observations or theory. The way models are currently calibrated is not a coincidence, % is coincidence the right word? 
but rather an inappropriate way to accommodate their complexity, where many parameters compensate for one another.

These limitations are central to current debates. In climate modeling, increasing model complexity has not necessarily led to reduced uncertainty. For instance, the uncertainty range on the effect of increasing CO2 concentration on temperature have remained largely unchanged \citep{Zelinka2020}. This has driven calls for more rigorous and transparent calibration processes \citep{balaji2022general}. Similar concerns arise in ecology, where strong discrepancies exist between model projections \citep{Cheaib2012}. Some researchers now advocate for the simplification of models, to avoid over-parametrization when the data provide little information to constrain some parameters \citep{Wang2017, Harrison2021}. If a model becomes too complex, understanding the sources of uncertainty and how they propagate through the model may become nearly impossible.
Each additional process and parameter can increase overall uncertainty to the point where predictions lose their usefulness for decision makers \citep{Saltelli2020}.
